---
title: "Absolute Risk Calculation: Black vs White" 
author: "Dezheng Huo" 
date: '2024-06-29'
fontsize: 9pt
output:
  html_document:
    df_print: paged
  word_document: default
header-includes: \usepackage{pdfpages}
geometry: margin=0.5in
fig_crop: no
---

```{r setup, include=FALSE}
#Load package
library(data.table)
library(readr)
library(dplyr)
library("tidyverse")
library(iCARE)
library(table1)
library(pROC)
library(ROCnReg)
library(readxl)
```
## [Method 2] Calculate absolute risk using iCare method
The life-time risk estimated from iCARE is based on PRS profile in validaiton set. It also rely on odds ratio. It does not matter if the PRS is calibrated or not. Family history profile is assumed controls from entire AABCG (9.4%).
### Functions defined
```{r}
#sigma <- qnorm(0.63)*sqrt(2)

set.seed(365895)

cat("Numbers with positive family hisotry", 1e6*0.094) 

pheno2 =  cbind.data.frame(case=c(rep(0, 1e6)), PRS=rnorm(1e6, mean=0, sd=1), 
                     famhis2=c(rep(0, 906000), rep(1, 94000))) 

## define the function of life time risk calculation
## define the function of life time risk calculation
calculate_LT_risk <- function(bc_inc, mort_inc, startAge, interval, OR1, OR2) {
  
  sublist1<-list(name="PRS",type="continuous")
  sublist2<-list(name="famhis2",type="factor",levels=c('0','1'),ref='0')

  new_cov_prof1 <- pheno2 %>% select(PRS, famhis2)
  head(new_cov_prof1)
  ref_cov_dat1 <- pheno2 %>% filter(case==0) %>% select(PRS, famhis2)
  bc_model_formula <- case ~ PRS + as.factor(famhis2)

  bc_model_cov_info <-list(sublist1, sublist2)
  bc_model_log_or <- c(log(OR1), log(OR2)) 
  names(bc_model_log_or) = c('PRS', 'as.factor(famhis2)1')
  bc_model_log_or

  res_covs_snps = computeAbsoluteRisk(model.formula=bc_model_formula, 
                                    model.cov.info=bc_model_cov_info, 
                                    model.log.RR=bc_model_log_or, 
                                    model.ref.dataset=ref_cov_dat1, 
                                    model.disease.incidence.rates=bc_inc,
                                    model.competing.incidence.rates=mort_inc,
                                    apply.age.start=startAge, 
                                    apply.age.interval.length=interval,
                                    apply.cov.profile=new_cov_prof1, 
                                    return.refs.risk=TRUE)

summary(res_covs_snps)
#plot(res_covs_snps, main="20-year risk: PRS + family history")
#head(res_covs_snps$details, 3)

  new_cov_auc <- cbind(pheno2, risk80B = as.numeric(res_covs_snps$risk))
  
  print("Model with PRS + famhis2")
  print("Correlation between the two risk scores")
  print(cor(new_cov_auc[,c(2:4)])) 
  
  return(new_cov_auc)
}

## Define the function to calcuate average risk per PRS percentiles.
 calculate_perc <- function(data, perc.u){

  perc.v <- c(perc.u[-1], 1) 
  prs_cat <- paste0(perc.u*100, "-", perc.v*100, "%") 
  data <- data[data$case==0, ]
  
  risk.percA <- quantile(data$PRS, c(perc.u, 1)) 
  print(round(risk.percA, digits=5)) 
  data$PRS.c <- cut(data$PRS, breaks=risk.percA, right=T, include.lowest=T, labels=prs_cat)

  risk.meanA <- as.data.frame(data %>% 
      group_by(PRS.c) %>%
      summarise_at(vars(risk80B), list(Total = mean)) ) 
    
  risk.meanB <- as.data.frame(data %>% filter(famhis2==1) %>% 
      group_by(PRS.c) %>%
      summarise_at(vars(risk80B), list(FamPos = mean)) ) 

  risk.meanC <- as.data.frame(data %>% filter(famhis2==0) %>% 
      group_by(PRS.c) %>%
      summarise_at(vars(risk80B), list(FamNeg = mean)) ) 

  risk.mean <- cbind(risk.meanA, risk.meanB, risk.meanC)
  risk.mean <- risk.mean[,c(1,2,4,6)]
  print(risk.mean)
  return(risk.mean)
 }
 
```
### BC incidence from SEER & mortality from other diseases from CDC
The incidence of BC are from from SEER22 2017-2021 data
Suggested Citation: 
SEER*Explorer: An interactive website for SEER cancer statistics [Internet]. Surveillance Research Program, National Cancer Institute; 2024 Apr 17. [updated: 2024 Jun 27; cited 2024 Jun 29]. Available from: https://seer.cancer.gov/statistics-network/explorer/. Data source(s): SEER Incidence Data, November 2023 Submission (1975-2021), SEER 22 registries.

The mortality rates are from CDC NCHS WONDER, 2017-2020
```{r}
## import data (data downloaded from SEER and CDC, reorganized in excel)
age_group_data_Black <- read_excel("G:/BCAC/james.li/PRS_AABCG/code/rscripts/DH_code/BreastCancerIncidence2021.xlsx")
age_black <- cbind(age_group_data_Black[, c(1,2)],
                   mort=age_group_data_Black$Total_mortality-age_group_data_Black$BC_mortality) 
age_black1 <- age_black[rep(row.names(age_black), each = 5), ]
age_black1[, 2:3] <- age_black1[, 2:3]/1e5

age_black_ERp <- cbind(age_group_data_Black[, c(1,6)],
                   mort=age_group_data_Black$Total_mortality-age_group_data_Black$BC_mortality) 
age_black_ERp1 <- age_black_ERp[rep(row.names(age_black_ERp), each = 5), ]
age_black_ERp1[, 2:3] <- age_black_ERp1[, 2:3]/1e5

age_black_ERn <- cbind(age_group_data_Black[, c(1,7)],
                   mort=age_group_data_Black$Total_mortality-age_group_data_Black$BC_mortality) 
age_black_ERn1 <- age_black_ERn[rep(row.names(age_black_ERn), each = 5), ]
age_black_ERn1[, 2:3] <- age_black_ERn1[, 2:3]/1e5

age_black_TNBC <- cbind(age_group_data_Black[, c(1,8)],
                   mort=age_group_data_Black$Total_mortality-age_group_data_Black$BC_mortality) 
age_black_TNBC1 <- age_black_TNBC[rep(row.names(age_black_TNBC), each = 5), ]
age_black_TNBC1[, 2:3] <- age_black_TNBC1[, 2:3]/1e5

age_group_data <- read_excel("G:/BCAC/james.li/PRS_AABCG/code/rscripts/DH_code/Incidence_otherRaces.xlsx")
names(age_group_data)
#age_group_data  ## per 100,000
age_white <- cbind(age_group_data[, 1:2], 
                   mort=age_group_data$Total_mortality_White-age_group_data$BC_mortality_White) 
age_white1 <- age_white[rep(row.names(age_white), each = 5), ]
age_white1[, 2:3] <- age_white1[, 2:3]/1e5

age_white_ERp <- cbind(age_group_data[, c(1,11)],
                    mort=age_group_data$Total_mortality_White-age_group_data$BC_mortality_White) 
age_white_ERp1 <- age_white_ERp[rep(row.names(age_white_ERp), each = 5), ]
age_white_ERp1[, 2:3] <- age_white_ERp1[, 2:3]/1e5

age_white_ERn <- cbind(age_group_data[, c(1,12)],
                    mort=age_group_data$Total_mortality_White-age_group_data$BC_mortality_White) 
age_white_ERn1 <- age_white_ERn[rep(row.names(age_white_ERn), each = 5), ]
age_white_ERn1[, 2:3] <- age_white_ERn1[, 2:3]/1e5 

age_white_TNBC <- cbind(age_group_data[, c(1,13)],
                    mort=age_group_data$Total_mortality_White-age_group_data$BC_mortality_White) 
age_white_TNBC1 <- age_white_TNBC[rep(row.names(age_white_TNBC), each = 5), ]
age_white_TNBC1[, 2:3] <- age_white_TNBC[, 2:3]/1e5


```
### Overall BC in Black and White
```{r}
## Blacks
age_black 
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_black1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_black1[,3]) )

result_OV <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=20, interval=60, OR1=1.488738, OR2=1)
perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_OV <- calculate_perc(data=result_OV, perc.u=perc.low) 

## White
age_white
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_white1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_white1[,3]) )

result_OV_white <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=20, interval=60, OR1=1.61, OR2=1)
perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_OV_white <- calculate_perc(data=result_OV_white, perc.u=perc.low) 

## Comparison of Black and White
  cbind.data.frame(PRS.c=perc_OV$PRS.c, Black=perc_OV[, c(2)]*100, White=perc_OV_white$Total*100)
 
  data1 <- rbind.data.frame(cbind.data.frame(Race="African American", risk80B=result_OV$risk80B), 
                            cbind.data.frame(Race="European American", risk80B=result_OV_white$risk80B))
  table1(~ risk80B | Race, data=data1, caption="Overall BC risk by race: Age 20-80", overall=F)

table(data1$risk80B>0.45, data1$Race)  
data1$risk80 <- data1$risk80B*100 
data1$risk80[data1$risk80>45] <- 45

 ggplot(data1, aes(x = risk80, fill = Race)) +
  geom_density(alpha = 0.4) +
  labs(
       x = "Lifetime risk of overall breast cancer (%)",
       y = "Density", 
       fill="Race") +
  scale_x_continuous(breaks = seq(0, 40, by = 5)) +
  theme_minimal() + theme_classic() + theme(legend.position = c(.80, .85))
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_BlackWhite_LT_OVERALL.png", width = 5, height = 3.5, dpi = 1200)
 
``` 
### ER+ BC in Black and White
```{r}
## Blacks
age_black_ERp 
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_black_ERp1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_black_ERp1[,3]) )

result_ERp <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=20, interval=60, OR1=1.526953, OR2=1)
perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_ERp <- calculate_perc(data=result_ERp, perc.u=perc.low) 

## White
age_white_ERp
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_white_ERp1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_white_ERp1[,3]) )

result_ERp_white <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=20, interval=60, OR1=1.68, OR2=1)
perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_ERp_white <- calculate_perc(data=result_ERp_white, perc.u=perc.low) 

## Comparison of Black and White
 
  cbind.data.frame(PRS.c=perc_ERp$PRS.c, Black=perc_ERp[, c(2)]*100, White=perc_ERp_white$Total*100)
 
  data1 <- rbind.data.frame(cbind.data.frame(Race="African American", risk80B=result_ERp$risk80B), 
                            cbind.data.frame(Race="European American", risk80B=result_ERp_white$risk80B))
  table1(~ risk80B | Race, data=data1, caption="ER+ BC risk by race: Age 20-80", overall=F)

table(data1$risk80B>0.40, data1$Race)  
data1$risk80 <- data1$risk80B*100 
data1$risk80[data1$risk80>40] <- 40

 ggplot(data1, aes(x = risk80, fill = Race)) +
  geom_density(alpha = 0.4) +
  labs(
       x = "Lifetime risk of ER-positive breast cancer (%)",
       y = "Density", 
       fill="Race") +
  scale_x_continuous(breaks = seq(0, 40, by = 5)) +
  theme_minimal() + theme_classic() + theme(legend.position = c(.80, .85))
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_BlackWhite_LT_ERPOS.png", width = 5, height = 3.5, dpi = 1200)
 
```
### ER- BC in Black and White
```{r}
## Blacks
age_black_ERn 
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_black_ERn1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_black_ERn1[,3]) )

result_ERn <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=20, interval=60, OR1=1.491039, OR2=1)
perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_ERn <- calculate_perc(data=result_ERn, perc.u=perc.low) 

## White
age_white_ERn
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_white_ERn1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_white_ERn1[,3]) )

result_ERn_white <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=20, interval=60, OR1=1.45, OR2=1)
perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_ERn_white <- calculate_perc(data=result_ERn_white, perc.u=perc.low) 

## Comparison of Black and White
 
  cbind.data.frame(PRS.c=perc_ERn$PRS.c, Black=perc_ERn[, c(2)]*100, White=perc_ERn_white$Total*100)
 
  data1 <- rbind.data.frame(cbind.data.frame(Race="African American", risk80B=result_ERn$risk80B), 
                            cbind.data.frame(Race="European American", risk80B=result_ERn_white$risk80B))
  table1(~ risk80B | Race, data=data1, caption="ER- BC risk by race: Age 20-80", overall=F)

table(data1$risk80B>0.12, data1$Race)  
data1$risk80 <- data1$risk80B*100 
data1$risk80[data1$risk80>12] <- 12

 ggplot(data1, aes(x = risk80, fill = Race)) +
  geom_density(alpha = 0.4) +
  labs(
       x = "Lifetime risk of ER-negative breast cancer (%)",
       y = "Density", 
       fill="Race") +
  scale_x_continuous(breaks = seq(0, 12, by = 2)) +
  theme_minimal() + theme_classic() + theme(legend.position = c(.80, .85))
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_BlackWhite_LT_ERNEG.png", width = 5, height = 3.5, dpi = 1200)
 
``` 
 
### TNBC in Black and White
```{r}
## Blacks
age_black_TNBC 
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_black_TNBC1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_black_TNBC1[,3]) )

result_TNBC <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=20, interval=60, OR1=1.653526, OR2=1)
perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_TNBC <- calculate_perc(data=result_TNBC, perc.u=perc.low) 

## White
age_white_TNBC
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_white_TNBC1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_white_TNBC1[,3]) )

result_TNBC_white <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=20, interval=60, OR1=1.65, OR2=1)
perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_TNBC_white <- calculate_perc(data=result_TNBC_white, perc.u=perc.low) 

## Comparison of Black and White
 
  cbind.data.frame(PRS.c=perc_TNBC$PRS.c, Black=perc_TNBC[, c(2)]*100, White=perc_TNBC_white$Total*100)
 
  data1 <- rbind.data.frame(cbind.data.frame(Race="African American", risk80B=result_TNBC$risk80B), 
                            cbind.data.frame(Race="European American", risk80B=result_TNBC_white$risk80B))
  table1(~ risk80B | Race, data=data1, caption="TNBC risk by race: Age 20-80", overall=F)

table(data1$risk80B>0.10, data1$Race)  
data1$risk80 <- data1$risk80B*100 
data1$risk80[data1$risk80>10] <- 10

 ggplot(data1, aes(x = risk80, fill = Race)) +
  geom_density(alpha = 0.4) +
  labs(
       x = "Lifetime risk of triple negative breast cancer (%)",
       y = "Density", 
       fill="Race") +
  scale_x_continuous(breaks = seq(0, 10, by = 2)) +
  theme_minimal() + theme_classic() + theme(legend.position = c(.80, .85))
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_BlackWhite_LT_TNBC.png", width = 5, height = 3.5, dpi = 1200)
 
``` 

## [Method 1] Absolute Risk Calculation, theoritical methods
To calculate 10-year risk for a 40, 45, and 50 years old AA. 
Base on ACS screening guideline, mammogram is recommended to start at age 45, so the average 10-year risk for a 45 years old can be a threshold for screening. 
It was 2.1% for African Americans and 2.3% for European Americans. We may take 2.2% for the entire population. 

### [Step 1] Define the function.
```{r}
calculate_lifetime_risk <- function(sigma, ShortTime, rate_inc, rate_mor, Ystart, perc.u, perc.v) {
  
  # Calculate PRS frequency
  prs_freq <- perc.v - perc.u 
  
  # Calculate sigma
  print(sigma)
  
  # Calculate odds ratio
  OR <- (0.6 - 0.4) * (pnorm(qnorm(1 - perc.u) + sigma) - pnorm(qnorm(1 - perc.v) + sigma)) / 
    (perc.v - perc.u) / (pnorm(qnorm(0.6) + sigma) - pnorm(qnorm(0.4) + sigma)) 
  
  # Create a data frame for odds ratio
  ORstring <- data.frame(
    g = 1:length(OR),
    prs_cat = paste0(perc.u * 100, "-", perc.v * 100, "%"),
    prs_freq,
    OR
  )
  
  # Print odds ratio per percentiles
  cat("Odds ratio per percentiles:\n")
  print(ORstring[, 2:4])
  
  Row <- length(OR)
  Col <- 81 - Ystart
  
  # Initialize matrices
  S_g <- matrix(, nrow = Row, ncol = Col)  ## BC-free matrix; row for category, column for age
  S_g[, 1] <- rep(1, Row) 
  AR_g_adj <- matrix(, nrow = Row, ncol = Col)  ## Cumulative risk matrix adjusted for competing risk
  AR_g_adj[, 1] <- rep(0, Row)
  # Cumulative risk matrix without competing risk, by the end of time period and the beginning of time period t+1. 
  AR <- matrix(, nrow = Row, ncol = Col)  ##  consideration
  AR[, 1] <- rep(0, Row)
  mu_0 <- numeric(Col)  ## Breast cancer hazard of baseline PRS category for each age
  S_m <- matrix(, nrow = Row, ncol = Col)  ## Survival from other diseases
  mortality <- matrix(, nrow = Row, ncol = Col) ## Mortality from other diseases
  S_m[, 1] <- rep(1, Row) 
  
  # Calculate risk matrices
  for (t in 1:(Col - 1)) {
    i_t <- rate_inc[t + Ystart - 20]   ## BC incidence
    mu_0[t] <- solve(sum(prs_freq * OR * S_g[, t]) / sum(prs_freq * S_g[, t]), i_t)
    S_g[, t + 1] <- S_g[, t] * (1 - mu_0[t] * OR)
    mortality[, t] <- rep(rate_mor[t + Ystart - 20], Row) 
    S_m[, t + 1] <- S_m[, t] * (1 - mortality[, t])
  }
  
  for (t in 1:(Col - 1)) {
    AR_g_adj[, t + 1] <- mu_0[t] * OR * S_g[, t] * S_m[, t] + AR_g_adj[, t] 
    AR[, t + 1] <- mu_0[t] * OR * S_g[, t] + AR[, t]  
  }
  
  rownames(AR) <- ORstring$prs_cat
  colnames(AR) <- Ystart:80
  AR_t <- t(AR * 100)
  cat(paste0("Cumulative risk from age ", Ystart, " to 80, without competing risk consideration: \n"))
  print(AR_t[nrow(AR_t), ], digits = 3)
  
  rownames(AR_g_adj) <- ORstring$prs_cat
  colnames(AR_g_adj) <- Ystart:80
  LT_risk <- t(AR_g_adj * 100)
  cat(paste0("Cumulative risk from age ", Ystart, " to 80, considering competing risk: \n"))
  print(LT_risk[nrow(LT_risk), ], digits = 3)
  
  # Calculate 5-year risk
  AR_ten <- matrix(, nrow = Row, ncol = (Col - ShortTime))
  for (t in 1:(Col - ShortTime)) {
    AR_ten[, t] <- (AR_g_adj[, t + ShortTime] - AR_g_adj[, t]) / (S_g[, t] * S_m[, t])
  }
  rownames(AR_ten) <- ORstring$prs_cat
  colnames(AR_ten) <- Ystart:(80 - ShortTime)
  Risk_5y <- t(AR_ten * 100)
  cat(paste0(ShortTime, "-year risk for women aged ", Ystart, ": \n")) 
  print(Risk_5y[1, ], digits = 3)
  
  # Combine results into a list
  risk_combine <- list(LT_risk, Risk_5y) 
  return(risk_combine)  
}

```

### [Step 2] Risk Calculation
#### Overall cancer risk, only based on age, not PRS 
```{r}
## Black
perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc.high <- c(perc.low[-1], 1)
  # qnorm(0.626) * sqrt(2)
  # 0.48567

result_black <- calculate_lifetime_risk(sigma=0, Ystart=20, ShortTime=10,
                                        rate_inc= age_black1[,2], rate_mor= age_black1[,3],  
                                        perc.u= perc.low, perc.v= perc.high)
print("Cumulative risk from age 20 to 80 is the same across PRS category as OR=1") 
as.data.frame(result_black[1])[61,] 
print("10 year risk for women ages 40, 45, and 50") 
 out.black <- cbind(Age=as.numeric(row.names(as.data.frame(result_black[2]))), as.data.frame(result_black[2])) 
 out.black[c(21,26,31), c(1:2)] 
 

## White
result_white <- calculate_lifetime_risk(sigma=0, Ystart=20, ShortTime=10,
                                        rate_inc= age_white1[,2], rate_mor= age_white1[,3],  
                                        perc.u= perc.low, perc.v= perc.high)
print("Cumulative risk from age 20 to 80 is the same across PRS category as OR=1") 
as.data.frame(result_white[1])[61,] 
print("10 year risk for women ages 40, 45, and 50") 
 out.white <- cbind(Age=as.numeric(row.names(as.data.frame(result_white[2]))), as.data.frame(result_white[2])) 
 out.white[c(21,26,31), c(1:2)] 
