---
title: "Absolute Risk Calculation: PRS + Family History" 
author: "Dezheng Huo" 
date: '2024-06-28'
fontsize: 9pt
output:
  html_document:
    df_print: paged
  word_document: default
header-includes: \usepackage{pdfpages}
geometry: margin=0.5in
fig_crop: no
---

```{r setup, include=FALSE}
#Load package
library(data.table)
library(readr)
library(dplyr)
library("tidyverse")
library(iCARE)
library(table1)
library(pROC)
library(ROCnReg)
library(readxl)
```
## [Method 2] Calculate absolute risk using iCare method
The life-time risk estimated from iCARE is based on PRS profile in validaiton set. It also rely on odds ratio. It does not matter if the PRS is calibrated or not. Family history profile is assumed controls from entire AABCG (9.4%).
### Functions defined
```{r}
#sigma <- qnorm(0.63)*sqrt(2)

set.seed(365895)

cat("Numbers with positive family hisotry", 1e6*0.094) 

pheno2 =  cbind.data.frame(case=c(rep(0, 1e6)), PRS=rnorm(1e6, mean=0, sd=1), 
                     famhis2=c(rep(0, 906000), rep(1, 94000))) 

## define the function of life time risk calculation
calculate_LT_risk <- function(bc_inc, mort_inc, startAge, interval, OR1, OR2) {
  
  sublist1<-list(name="PRS",type="continuous")
  sublist2<-list(name="famhis2",type="factor",levels=c('0','1'),ref='0')

  new_cov_prof1 <- pheno2 %>% select(PRS, famhis2)
  head(new_cov_prof1)
  ref_cov_dat1 <- pheno2 %>% filter(case==0) %>% select(PRS, famhis2)
  bc_model_formula <- case ~ PRS + as.factor(famhis2)

  bc_model_cov_info <-list(sublist1, sublist2)
  bc_model_log_or <- c(log(OR1), log(OR2)) 
  names(bc_model_log_or) = c('PRS', 'as.factor(famhis2)1')
  bc_model_log_or

  res_covs_snps = computeAbsoluteRisk(model.formula=bc_model_formula, 
                                    model.cov.info=bc_model_cov_info, 
                                    model.log.RR=bc_model_log_or, 
                                    model.ref.dataset=ref_cov_dat1, 
                                    model.disease.incidence.rates=bc_inc,
                                    model.competing.incidence.rates=mort_inc,
                                    apply.age.start=startAge, 
                                    apply.age.interval.length=interval,
                                    apply.cov.profile=new_cov_prof1, 
                                    return.refs.risk=TRUE)

summary(res_covs_snps)
#plot(res_covs_snps, main="20-year risk: PRS + family history")
#head(res_covs_snps$details, 3)

  new_cov_auc <- cbind(pheno2, risk80B = as.numeric(res_covs_snps$risk))
  
  print("Model with PRS + famhis2")
  print("Correlation between the two risk scores")
  print(cor(new_cov_auc[,c(2:4)])) 
  
  return(new_cov_auc)
}


## Define the function of risk cutoff
 calculate_perc <- function(data, perc.u){

  perc.v <- c(perc.u[-1], 1) 
  prs_cat <- paste0(perc.u*100, "-", perc.v*100, "%") 
  data <- data[data$case==0, ]
  
  risk.percA <- quantile(data$PRS, c(perc.u, 1)) 
  print(round(risk.percA, digits=5)) 
  data$PRS.c <- cut(data$PRS, breaks=risk.percA, right=T, include.lowest=T, labels=prs_cat)

  risk.meanB <- as.data.frame(data %>% filter(famhis2==1) %>% 
      group_by(PRS.c) %>%
      summarise_at(vars(risk80B), list(FamPos = mean)) ) 

  risk.meanC <- as.data.frame(data %>% filter(famhis2==0) %>% 
      group_by(PRS.c) %>%
      summarise_at(vars(risk80B), list(FamNeg = mean)) ) 

  risk.mean <- cbind(risk.meanB, risk.meanC)
  risk.mean <- risk.mean[,c(1,2,4)]
  print(risk.mean)
  return(risk.mean)
 }
 
```
### BC incidence from SEER & mortality from other diseases from CDC
The incidence of BC are from from SEER22 2017-2021 data
The mortality rates are from CDC NCHS WONDER, 2017-2020
```{r}
## import data (data downloaded from SEER and CDC, reorganized in excel)
age_group_data_Black <- read_excel("G:/BCAC/james.li/PRS_AABCG/code/rscripts/DH_code/BreastCancerIncidence2021.xlsx")
age_black <- cbind(age_group_data_Black[, c(1,2)],
                   mort=age_group_data_Black$Total_mortality-age_group_data_Black$BC_mortality) 
age_black1 <- age_black[rep(row.names(age_black), each = 5), ]
age_black1[, 2:3] <- age_black1[, 2:3]/1e5

age_black_ERp <- cbind(age_group_data_Black[, c(1,6)],
                   mort=age_group_data_Black$Total_mortality-age_group_data_Black$BC_mortality) 
age_black_ERp1 <- age_black_ERp[rep(row.names(age_black_ERp), each = 5), ]
age_black_ERp1[, 2:3] <- age_black_ERp1[, 2:3]/1e5

age_black_ERn <- cbind(age_group_data_Black[, c(1,7)],
                   mort=age_group_data_Black$Total_mortality-age_group_data_Black$BC_mortality) 
age_black_ERn1 <- age_black_ERn[rep(row.names(age_black_ERn), each = 5), ]
age_black_ERn1[, 2:3] <- age_black_ERn1[, 2:3]/1e5

age_black_TNBC <- cbind(age_group_data_Black[, c(1,8)],
                   mort=age_group_data_Black$Total_mortality-age_group_data_Black$BC_mortality) 
age_black_TNBC1 <- age_black_TNBC[rep(row.names(age_black_TNBC), each = 5), ]
age_black_TNBC1[, 2:3] <- age_black_TNBC1[, 2:3]/1e5

```
### Risk Calculation: Overall BC
```{r}
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_black1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_black1[,3]) )

result_OV <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=20, interval=60, OR1=1.509951, OR2=3.058138)
 
result_OV <- result_OV %>% mutate(group= case_when (famhis2 == 0 ~ "No",
                                                 famhis2 == 1 ~ "Yes"))
table1(~ risk80B | group, data=result_OV, caption="Overall BC risk by family history: Age 20-80")

perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_OV <- calculate_perc(data=result_OV, perc.u=perc.low) 

table(result_OV$risk80B>0.5, result_OV$famhis2)  
result_OV$risk80B[result_OV$risk80B>0.5] <- 0.5 
result_OV$risk80 <- result_OV$risk80B*100

 ggplot(result_OV, aes(x = risk80, fill = group)) +
  geom_density(aes(y = after_stat(count)/sum(count)), alpha = 0.4) +
  labs(title = "  Overall breast cancer",
       x = "Lifetime risk (%)",
       y = "Density", 
       fill="Family History") +
  scale_x_continuous(breaks = seq(0, 50, by = 10)) +
  theme_minimal() + theme_classic() + theme(legend.position = c(.85, .85))
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_LT_OVERALL.png", width = 5, height = 3.5, dpi = 1200)
 

result_OV2 <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=40, interval=10, OR1=1.509951, OR2=3.058138)
 
result_OV2 <- result_OV2 %>% mutate(group= case_when (famhis2 == 0 ~ "No",
                                                 famhis2 == 1 ~ "Yes"))
table1(~ risk80B | group, data=result_OV2, caption="Overall BC risk by family history: Age 40-50")

perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_OV2 <- calculate_perc(data=result_OV2, perc.u=perc.low) 

table(result_OV2$risk80B>0.08, result_OV2$famhis2)  
result_OV2$risk80B[result_OV2$risk80B>0.08] <- 0.08
result_OV2$risk80 <- result_OV2$risk80B*100

 ggplot(result_OV2, aes(x = risk80, fill = group)) +
  geom_density(aes(y = after_stat(count)/sum(count)), alpha = 0.4) +
  labs(title = "  Overall breast cancer",
       x = "10-year risk for women at age 40 (%)",
       y = "Density", 
       fill="Family History") +
  scale_x_continuous(breaks = seq(0, 8, by = 1)) +
    theme_minimal() + theme_classic() + theme(legend.position = c(.85, .85)) 
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_age40_10yr_OVERALL.png", width = 5, height = 3.5, dpi = 1200)

 out_OV <- cbind(perc_OV, perc_OV2[, 2:3])
 colnames(out_OV) <- c('PRS.c', 'LT_FamPos', 'LT_FamNeg', 'y10_FamPos', 'y10_FamNeg') 
 write.csv(out_OV, "G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Risk_FamilyHistory_OVERALL.csv", row.names = F)
 
``` 

### ER+ BC 
```{r}
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_black_ERp1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_black_ERp1[,3]) )

result_ERp <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, OR1=1.557937, OR2=3.014389, startAge=20, interval=60)
 
result_ERp <- result_ERp %>% mutate(group= case_when (famhis2 == 0 ~ "No",
                                                 famhis2 == 1 ~ "Yes"))
table1(~ risk80B | group, data=result_ERp, caption="ER+ BC risk by family history: Age 20-80")

perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_ERp <- calculate_perc(data=result_ERp, perc.u=perc.low) 

table(result_ERp$risk80B>0.45, result_ERp$famhis2)  
result_ERp$risk80B[result_ERp$risk80B>0.45] <- 0.45 
result_ERp$risk80 <- result_ERp$risk80B*100

 ggplot(result_ERp, aes(x = risk80, fill = group)) +
  geom_density(aes(y = after_stat(count)/sum(count)), alpha = 0.4) +
  labs(title = "  ER+ breast cancer",
       x = "Lifetime risk (%)",
       y = "Density", 
       fill="Family History") +
  scale_x_continuous(breaks = seq(0, 50, by = 10)) +
    theme_minimal() + theme_classic() + theme(legend.position = c(.85, .85)) 
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_LT_ERPOS.png", width = 5, height = 3.5, dpi = 1200)
 

result_ERp2 <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, startAge=40, interval=10, OR1=1.557937, OR2=3.014389)
 
result_ERp2 <- result_ERp2 %>% mutate(group= case_when (famhis2 == 0 ~ "No",
                                                 famhis2 == 1 ~ "Yes"))
table1(~ risk80B | group, data=result_ERp2, caption="ER+ BC risk by family history: Age 40-50")

perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_ERp2 <- calculate_perc(data=result_ERp2, perc.u=perc.low) 

table(result_ERp2$risk80B>0.075, result_ERp2$famhis2)  
result_ERp2$risk80B[result_ERp2$risk80B>0.075] <- 0.075
result_ERp2$risk80 <- result_ERp2$risk80B*100

 ggplot(result_ERp2, aes(x = risk80, fill = group)) +
  geom_density(aes(y = after_stat(count)/sum(count)), alpha = 0.4) +
  labs(title = "  ER+ breast cancer",
       x = "10-year risk for women at age 40 (%)",
       y = "Density", 
       fill="Family History") +
  scale_x_continuous(breaks = seq(0, 7, by = 1)) +
    theme_minimal() + theme_classic() + theme(legend.position = c(.85, .85)) 
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_age40_10yr_ERPOS.png", width = 5, height = 3.5, dpi = 1200)
 
 out_ERp <- cbind(perc_ERp, perc_ERp2[, 2:3])
 colnames(out_ERp) <- c('PRS.c', 'LT_FamPos', 'LT_FamNeg', 'y10_FamPos', 'y10_FamNeg') 
 write.csv(out_ERp, "G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Risk_FamilyHistory_ERPOS.csv", row.names = F)
 
``` 
### ER- BC 
```{r}
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_black_ERn1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_black_ERn1[,3]) )

result_ERn <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, OR1=1.455197, OR2=2.641937, startAge=20, interval=60)
 
result_ERn <- result_ERn %>% mutate(group= case_when (famhis2 == 0 ~ "No",
                                                 famhis2 == 1 ~ "Yes"))
table1(~ risk80B | group, data=result_ERn, caption="ER+ BC risk by family history: Age 20-80")

perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_ERn <- calculate_perc(data=result_ERn, perc.u=perc.low) 

table(result_ERn$risk80B>0.18, result_ERn$famhis2)  
result_ERn$risk80B[result_ERn$risk80B>0.18] <- 0.18 
result_ERn$risk80 <- result_ERn$risk80B*100

 ggplot(result_ERn, aes(x = risk80, fill = group)) +
  geom_density(aes(y = after_stat(count)/sum(count)), alpha = 0.4) +
  labs(title = "  ER- breast cancer",
       x = "Lifetime risk (%)",
       y = "Density", 
       fill="Family History") +
  scale_x_continuous(breaks = seq(0, 15, by = 5)) +
    theme_minimal() + theme_classic() + theme(legend.position = c(.85, .85)) 
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_LT_ERNEG.png", width = 5, height = 3.5, dpi = 1200)
 
 
result_ERn2 <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, OR1=1.455197, OR2=2.641937, startAge=40, interval=10)
 
result_ERn2 <- result_ERn2 %>% mutate(group= case_when (famhis2 == 0 ~ "No",
                                                 famhis2 == 1 ~ "Yes"))
table1(~ risk80B | group, data=result_ERn2, caption="ER- BC risk by family history: Age 40-50")

perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_ERn2 <- calculate_perc(data=result_ERn2, perc.u=perc.low) 

table(result_ERn2$risk80B>0.027, result_ERn2$famhis2)  
result_ERn2$risk80B[result_ERn2$risk80B>0.027] <- 0.027
result_ERn2$risk80 <- result_ERn2$risk80B*100

 ggplot(result_ERn2, aes(x = risk80, fill = group)) +
  geom_density(aes(y = after_stat(count)/sum(count)), alpha = 0.4) +
  labs(title = "  ER- breast cancer",
       x = "10-year risk for women at age 40 (%)",
       y = "Density", 
       fill="Family History") +
  scale_x_continuous(breaks = seq(0, 2.5, by = 0.5)) +
    theme_minimal() + theme_classic() + theme(legend.position = c(.85, .85)) 
 
 ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_age40_10yr_ERNEG.png", width = 5, height = 3.5, dpi = 1200)
 
 out_ERn <- cbind(perc_ERn, perc_ERn2[, 2:3])
 colnames(out_ERn) <- c('PRS.c', 'LT_FamPos', 'LT_FamNeg', 'y10_FamPos', 'y10_FamNeg') 
 write.csv(out_ERn, "G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Risk_FamilyHistory_ERNEG.csv", row.names = F)
 
``` 

### TNBC 
```{r}
rate_inc <- cbind(c(0:89),  c(rep(0,20), age_black_TNBC1[,2]) )
rate_mort <- cbind(c(0:89), c(rep(0,20), age_black_TNBC1[,3]) )

result_TNBC <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, OR1=1.655554, OR2=2.879107, startAge=20, interval=60)

result_TNBC <- result_TNBC %>% mutate(group= case_when (famhis2 == 0 ~ "No",
                                                        famhis2 == 1 ~ "Yes"))
table1(~ risk80B | group, data=result_TNBC, caption="TNBC risk by family history: Age 20-80")

perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_TNBC <- calculate_perc(data=result_TNBC, perc.u=perc.low) 

table(result_TNBC$risk80B>0.12, result_TNBC$famhis2)  
result_TNBC$risk80B[result_TNBC$risk80B>0.12] <- 0.12 
result_TNBC$risk80 <- result_TNBC$risk80B*100

ggplot(result_TNBC, aes(x = risk80, fill = group)) +
  geom_density(aes(y = after_stat(count)/sum(count)), alpha = 0.4) +
  labs(title = "  Triple-negative breast cancer",
       x = "Lifetime risk (%)",
       y = "Density", 
       fill="Family History") +
  scale_x_continuous(breaks = seq(0, 12, by = 2)) +
  theme_minimal() + theme_classic() + theme(legend.position = c(.85, .85)) 

ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_LT_TNBC.png", width = 5, height = 3.5, dpi = 1200)


result_TNBC2 <- calculate_LT_risk(bc_inc= rate_inc, mort_inc=rate_mort, OR1=1.655554, OR2=2.879107, startAge=40, interval=10)

result_TNBC2 <- result_TNBC2 %>% mutate(group= case_when (famhis2 == 0 ~ "No",
                                                          famhis2 == 1 ~ "Yes"))
table1(~ risk80B | group, data=result_TNBC2, caption="TNBC risk by family history: Age 40-50")

perc.low <-c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99)/100
perc_TNBC2 <- calculate_perc(data=result_TNBC2, perc.u=perc.low) 

table(result_TNBC2$risk80B>0.022, result_TNBC2$famhis2)  
result_TNBC2$risk80B[result_TNBC2$risk80B>0.022] <- 0.022
result_TNBC2$risk80 <- result_TNBC2$risk80B*100

ggplot(result_TNBC2, aes(x = risk80, fill = group)) +
  geom_density(aes(y = after_stat(count)/sum(count)), alpha = 0.4) +
  labs(title = "  Triple-negative breast cancer",
       x = "10-year risk for women at age 40 (%)",
       y = "Density", 
       fill="Family History") +
  scale_x_continuous(breaks = seq(0, 2, by = 0.5)) +
  theme_minimal() + theme_classic() + theme(legend.position = c(.85, .85))

ggsave("G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Density_age40_10yr_TNBC.png", width = 5, height = 3.5, dpi = 1200)

out_TNBC <- cbind(perc_TNBC, perc_TNBC2[, 2:3])
colnames(out_TNBC) <- c('PRS.c', 'LT_FamPos', 'LT_FamNeg', 'y10_FamPos', 'y10_FamNeg') 
write.csv(out_TNBC, "G:/BCAC/james.li/PRS_AABCG/output/LT_RISK_OUTPUT/Risk_FamilyHistory_TNBC.csv", row.names = F)

 
```
